---
title: "How does the design matrix help us"
author: "Ian Dworkin"
date: "28/02/2022"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# How does the design matrix help?

Tyler asked a very sensible question in class today about how (since the design matrix only has predictor variables) helps us to get our estimates.

From reading in Irizarry and Love, you (and Tyler) probably saw this on pages 160-161 of that book


$$
\mathbf{Y} = \mathbf{X\beta + \epsilon}
$$
Where $\mathbf{Y}$ is our vector of response, $\mathbf{X}$ s the design matrix,  $\mathbf{\beta}$ is the vector of coefficients and  $\mathbf{\epsilon}$ are the residual errors.

$$
\mathbf{Y} = \mathbf{X\beta + \epsilon}
$$
Using least squares estimation we want to minimize the residuals, so in matrix form this looks like


$$
 \mathbf{\hat{Y}} = \mathbf{X\beta}
$$


$$ 
( \mathbf{Y}- \mathbf{X\beta})^T( \mathbf{Y}- \mathbf{X\beta})
$$
Where the fitted values $\mathbf{\hat{Y}}$ are given by $\mathbf{X\beta}$.

Some calculus, and setting the derivative to zero allows us to solve for the coefficients we want to estimate


$$
\mathbf{\beta} = (\mathbf{X^TX)^{-1}X^TY}
$$

(Please note I had the the last part of the equation in the reverse order on the board, which will not work. Apologies.)


## That is probably not so intuitive...

Let's take a step back. So think about a really simple made up example with just a single continuous predictor:

```{r}
x <- 1:20

y <- rnorm(length(x), 2 + 0.5*x, 1.5)

plot(y ~ x, pch = 20)
```


We can fit the regression and see what the slope is

```{r}
lm(y ~ x)
```


### The way you learned it undergrad

You probably remember in your undergraduate stats class calculating the slope of this relationship by something that looks like this:


$$
\beta_1 = \frac{cov(x,y)}{\sigma^2_x}
$$
where the slope $\beta_1 =$, is the covariance between our response *y* and predictor *x* ($cov(x,y)$), divided by the variance of the predictor variable, *x* ($\sigma^2_x$). In R this would look like:

```{r}
cov(x,y)/var(x)
```

Which gives us the same slope! This works well for this simple example. But how do we extend it when we have multipe predictors.

Well the design matrix (and the matrix algebra) allows us to generalize this idea.

In this case our design matrix mmX (for $\mathbf{X}$) looks like:

```{r}
mmX <- model.matrix(~x)
mmX 
```

and we can use a bit of matrix algebra from above in R:
(**DON'T** worry about the details, just showing that it can be done this way, you don't need to know it if you don't want to)

```{r}
solve(t(mmX) %*% mmX) %*% t(mmX) %*% y
```


## Connecting the dots.

So one way of thinking about this is connecting the numerator and denominator for the simple case with what the matrix algebra is generalizing.

In this case the the covariance between x and y $cov(x,y)$ is mostly in this piece $\mathbf{X^TY}$, while the variance of the predictor that you are dividing by is mostly from $\mathbf{X^TX}^{-1}$. 



